08_relative_N_inputs

### 2025-06-20 08_relative_N_inputs.qmd
Tidy script, save intermediates

---
title: "08_relative_N_inputs"
format: html
editor: source
---

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(here)
library(terra)
library(sf)
library(qs)
library(ggplot2)
library(tidyterra)
library(cowplot)
library(rnaturalearth)

@@

Create a blank raster with enough resolution to get farms in there (100m)

```{r base-raster}
base_rast <- rast(res = 0.1/111.32, extent = ext(-180, 180, -90, 90), crs = "EPSG:4326")

@@

inset_rasters <- list(
  CAN1 = crop(base_rast, ext(inset_boxes[["CAN1"]])),
  CAN2 = crop(base_rast, ext(inset_boxes[["CAN2"]])),
  EUR  = crop(base_rast, ext(inset_boxes[["EUR"]])),
  CHI  = crop(base_rast, ext(inset_boxes[["CHI"]])),
  AUS  = crop(base_rast, ext(inset_boxes[["AUS"]]))
)
```

@@

Data was downloaded from https://doi.org/10.5063/F1610XPS. Resolution is ~934 x 934 m in the Mollweide projection.

```{r raw-back-N}
back_N_rast <- file.path(input_Ndata_path, "nutrient_pollution_2013_raw.tif") %>% 
    rast()
back_N_rast
```

When extracting background N data, it was found that a lot of cells had no data. To try and fix this somewhat, I'm going to reproject the background N onto a slightly coarser grid.

```{r small-res-change}
back_N_rast <- back_N_rast %>% 
  project(rast(res = c(1000, 1000), extent = ext(back_N_rast), crs = crs(back_N_rast)),
          method  = "average")
back_N_rast
```

@@

```{r farm locations}
farm_locs <- file.path(output_farm_data_path, "farm_geometry.qs") %>% qs::qread()
farm_sfs <- st_sf(farm_id = farm_locs$farm_id, geometry = farm_locs$geometry) %>% 
  st_transform(crs = crs(back_N_rast))
farm_rast <- farm_sfs %>% 
  rasterize(back_N_rast, field = "farm_id")
farm_rast
```

>>>>

---
title: "08_relative_N_inputs"
author: Tormey Reimer
date: today
editor: source
---

# Introduction

The purpose of this markdown is to:

1. Import and clean up background nitrogen data
2. Overlay background nitrogen and Atlantic salmon farm locations
3. Compare annual inputs from the different sources

```{r setup}
#| code-summary: R setup

library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)
library(here)
library(terra)
library(sf)
library(qs)
library(ggplot2)
library(tidyterra)
library(cowplot)
library(rnaturalearth)
library(purrr)

@@

```{r base-raster}
#| code-summary: Create some blank rasters with parameters to be used going forward

@@

base_wgs84 <- rast(res = c(0.008333333, 0.008333333), extent = ext(c(-180, 180, -90, 90)), crs = "EPSG:4326")
base_molle <- rast(res = c(1000, 1000), extent = ext(c(-18040096, 18040096, -9020048, 9020048)), crs = "ESRI:54009")

wgs84_insets <- inset_boxes %>% 
  lapply(function(bx) {crop(base_wgs84, ext(bx))})

molle_insets <- wgs84_insets %>% 
  lapply(function(rs) {project(rs, crs(base_molle))}) %>% 
  lapply(function(rs) {ext(rs)}) %>% 
  lapply(function(ex) {crop(base_molle, ex)})
```

@@

Data for nutrient runoff from agriculture fertilisers and pesticides was downloaded from @halpern_recent_2019-1 [dataset here: @halpern_recent_2019].

```{r raw-back-N-halpern}
#| code-summary: Import data from @halpern_recent_2019 and crop/reproject for the 5 insets being used
#| eval: false

halpern_N_rast <- file.path(input_Ndata_path, "halpern_2019", "nutrient_pollution_2013_raw.tif") %>% rast()

halpern_insets <- molle_insets %>% 
  lapply(function(ins) {project(halpern_N_rast, ins, method = "cubic")})

# Change to the Molleweide projection
tuholske_N_rast <- file.path(input_Ndata_path, "tuholske_2021", "global_effluent_2015_tot_N.tif") %>% rast() 
tuholske_insets <- molle_insets %>% 
  lapply(function(ins) {project(tuholske_N_rast, ins, method = "cubic")})

# The halpern data has NA values for land but the tuholske data does not
tuholske_insets <- map2(tuholske_insets, halpern_insets, function(rast, mask) {
  mask(rast, mask)
})

background_N_insets <- map2(halpern_insets, tuholske_insets, function(dat1, dat2) {dat1 + dat2})

map2(names(halpern_insets), background_N_insets, function(nm, rast) {
  writeRaster(rast, file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")))
})
```

Data for inputs of human sewage in coastal ecosystems was sourced from @tuholske_global_2021 [dataset here: @tuholske_mapping_2021]. 
The data is in annual g N.

```{r raw-back-N-tuholske}
#| code-summary: Import data from @tuholske_global_2021 and crop/reproject for the 5 insets being used
#| eval: false

# Change to the Molleweide projection
tuholske_N_rast <- file.path(input_Ndata_path, "tuholske_2021", "global_effluent_2015_tot_N.tif") %>% rast() 
tuholske_insets <- molle_insets %>% 
  lapply(function(ins) {project(tuholske_N_rast, ins, method = "cubic")})

# The halpern data has NA values for land but the tuholske data does not
tuholske_insets <- map2(tuholske_insets, halpern_insets, function(rast, mask) {
  mask(rast, mask)
})
```

```{r save-back-N}
#| code-summary: Combine and save background N data

background_N_insets <- map2(
  halpern_insets, 
  tuholske_insets, 
  function(halpern, tuholske) {
    halpern + tuholske
  })

map2(names(halpern_insets), background_N_insets, function(nm, rast) {
  writeRaster(rast, file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")))
})
```

```{r load-background-data}
#| code-summary: Load previously saved inset rasters

background_N_insets <- lapply(names(inset_boxes), function(nm) {
  file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")) %>% rast()
})
```

@@

```{r farm locations}
#| code-summary: Bring in modelled salmon inputs data

farm_locs <- file.path(output_farm_data_path, "farm_geometry.qs") %>% qs::qread()
farm_sfs <- st_sf(farm_id = farm_locs$farm_id, geometry = farm_locs$geometry) %>% 
  st_transform(crs = crs(base_molle))
farm_id_insets <- background_N_insets %>% 
  lapply(function(ins) {rasterize(farm_sfs, ins, field = "farm_id")})
```
